---
title: "Logistic Regression and Further Study"
output: html_document
date: "2025-11-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r required libraries}
pkgs <- c("pROC", "car", "glmnet", "logistf")

for(p in pkgs){
    if (!require(p,character.only=TRUE)) install.packages(p)
}
```

```{r read}
file_path <- c("C:/Users/billy/DataScienceFanClub/Data/Val_traindata/df_train_log_scaled_val.csv", "C:/Users/billy/DataScienceFanClub/Data/Val_traindata/df_test_log_scaled_val.csv")
train_val <- read.csv(file_path[1])
test_val <- read.csv(file_path[2])
```

We will now convert the status year columns into numerical targets, a form of binary encoding, with 0 for "alive" and 1 for "failed".

```{r binary encoding}
cols <- c("status_label_1yr", "status_label_2yr", "status_label_3yr")

for (col in cols) {
  train_val[[col]] <- ifelse(train_val[[col]] == "failed", 1, 0)
  test_val[[col]]  <- ifelse(test_val[[col]] == "failed", 1, 0)
}
```

As explained in the group part of the report, we have introduced lag variables so it is important that we categorise them for easier use in the future.

```{r feature categorisation}
features <- paste0("X", 1:18)
features <- features[features !=  "X16"]
lag_1yr_feats <- paste0(features, "_1yr")
lag_2yr_feats <- paste0(features, "_2yr")
lag_3yr_feats <- paste0(features, "_3yr")

print(features)
```

We will now plot our simple logistic regression model, only considering base features and with respect to status_label_1yr.

```{r baseline test}
formula_base <- as.formula(paste("status_label_1yr ~", paste(features, collapse = " + ")))
model_base <- glm(formula_base, data = train_val, family = binomial())
summary(model_base)
```

Recall that from the EDA section that there is high correlation between variables. We will explore this more after our initial testing. 

```{r loop for models}
formulas <- list(
  base = as.formula(paste("status_label_1yr ~", paste(features, collapse = " + "))),
  base_lag1 = as.formula(paste("status_label_1yr ~", paste(c(features, lag_1yr_feats), collapse = " + "))),
  base_lag1_2 = as.formula(paste("status_label_1yr ~", paste(c(features, lag_1yr_feats, lag_2yr_feats), collapse = " + "))),
  all = as.formula(paste("status_label_1yr ~", paste(c(features, lag_1yr_feats, lag_2yr_feats, lag_3yr_feats), collapse = " + ")))
)

models <- list()
for (i in names(formulas)) {
  f <- formulas[[i]]
  models[[i]] <- glm(f, data = train_val, family = binomial())
}

for (i in names(models)) {
  cat("\nModel:", i, "\n")
  cat("AIC:", AIC(models[[i]]), "\n")
}
```

```{r testing vale}
glm_probs <- list()
for (i in names(models)) {
  mod <- models[[i]]
  glm_probs[[i]] <- predict(mod, newdata = test_val, type = "response")
}

roc_list <- list()
auc_values <- numeric()

for (i in names(glm_probs)) {
  roc_list[[i]] <- roc(test_val$status_label_1yr, glm_probs[[i]])
  auc_values[i] <- auc(roc_list[[i]])
}

# view AUC results
print(round(auc_values, 3))

plot(roc_list[[1]], col = 1, lwd = 2, main = "ROC Curves by Model")
for (j in 2:length(roc_list)) {
  lines(roc_list[[j]], col = j, lwd = 2)
}

legend("bottomright",
       legend = paste(names(roc_list), "AUC =", round(auc_values, 3)),
       col = 1:length(roc_list), lwd = 2, bty = "n")

auc_df <- data.frame(
  Model = c("Base", "Base + Lag1", "Base + Lag1 + Lag 2", "Base + All Lags"),
  AUC = c(auc_values["base"], auc_values["base_lag1"], auc_values["base_lag1_2"], auc_values["all"])
)

print(auc_df)

best_index <- which.max(auc_df$AUC)
best_model_name <- auc_df$Model[best_index]
print(best_model_name)

thresh <- coords(roc_list[[1]], "best", ret = "threshold")
glm.pred <- ifelse(glm_probs[[1]] > thresh$threshold, 1, 0)
actual <- test_val$status_label_1yr

cm <- table(Predicted = glm.pred, Actual = actual)
cm

TP <- cm[2, 2]  
TN <- cm[1, 1]  
FP <- cm[2, 1]  
FN <- cm[1, 2]

accuracy  <- (TP + TN) / sum(cm)
precision <- TP / (TP + FP)
recall    <- TP / (TP + FN)
f1_score  <- 2 * precision * recall / (precision + recall)

cat("Accuracy :", round(accuracy, 3), "\n")
cat("Precision:", round(precision, 3), "\n")
cat("Recall   :", round(recall, 3), "\n")
cat("F1 Score :", round(f1_score, 3), "\n")
```

We see that the the base model has the highest AUC. We then calculate the confusion matrix and see that we have low precision but high recall. This is because our model is incorrectly flagging many companies as bankrupt when they do in fact not go bankrupt. However, the high recall score does mean that the model is correctly flagging companies that are bankrupt as correct, which is the more important application. Thus, with a score of 0.9 our base model can correctly distinguish between bankrupt and non bankrupt companies very well.

It is important to note that even with a high AUC score it does not necessarily mean our model is necessarily excellent. This is because of the low precision score. Therefore, if one were to improve this whilst maintaining a high AUC, then it would lend credence into the belief that the model is truthfully "excellent".

We now extend our findings by removing all highly correlated variables. We choose said variables by the correlation analysis done in the EDA section of the report.

```{r removal variables}
new_features <- features[-c(2,4,9,10)]
lag_1yr_newfeats <- paste0(new_features, "_1yr")
lag_2yr_newfeats <- paste0(new_features, "_2yr")
lag_3yr_newfeats <- paste0(new_features, "_3yr")

print(new_features)
```

```{r new feat}
formulas_new <- list(
  base_new = as.formula(paste("status_label_1yr ~", paste(new_features, collapse = " + "))),
  base_lag1_new = as.formula(paste("status_label_1yr ~", paste(c(new_features, lag_1yr_newfeats), collapse = " + "))),
  base_lag1_2_new = as.formula(paste("status_label_1yr ~", paste(c(new_features, lag_1yr_newfeats, lag_2yr_newfeats), collapse = " + "))),
  all_new = as.formula(paste("status_label_1yr ~", paste(c(new_features, lag_1yr_newfeats, lag_2yr_newfeats, lag_3yr_newfeats), collapse = " + ")))
)

models_new <- list()
for (i in names(formulas_new)) {
  f <- formulas_new[[i]]
  models_new[[i]] <- glm(f, data = train_val, family = binomial())
}

for (i in names(models_new)) {
  cat("\nModel:", i, "\n")
  cat("AIC:", AIC(models_new[[i]]), "\n")
}
```

```{r roc curves on new features}
glm_probs_new <- list()
for (i in names(models_new)) {
  mod <- models_new[[i]]
  glm_probs_new[[i]] <- predict(mod, newdata = test_val, type = "response")
}

roc_list_new <- list()
auc_values_new <- numeric()

# loop through all predicted probability vectors
for (i in names(glm_probs_new)) {
  roc_list_new[[i]] <- roc(test_val$status_label_1yr, glm_probs_new[[i]])
  auc_values_new[i] <- auc(roc_list_new[[i]])
}

# view AUC results
print(round(auc_values_new, 3))

# list roc plot
plot(roc_list_new[[1]], col = 1, lwd = 2, main = "ROC Curves by Model")
for (j in 2:length(roc_list_new)) {
  lines(roc_list_new[[j]], col = j, lwd = 2)
}

legend("bottomright",
       legend = paste(names(roc_list_new), "AUC =", round(auc_values_new, 3)),
       col = 1:length(roc_list_new), lwd = 2, bty = "n")

auc_df_new <- data.frame(
  Model = c("Base", "Base + Lag1", "Base + Lag1 + Lag2", "Base + All Lags"),
  AUC = c(auc_values_new["base_new"], auc_values_new["base_lag1_new"], auc_values_new["base_lag1_2_new"], auc_values_new["all_new"])
)

print(auc_df_new)

best_index_new <- which.max(auc_df_new$AUC)
best_model_name_new <- auc_df_new$Model[best_index_new]
print(best_model_name_new)
```

```{r new accuracy}
thresh_new <- coords(roc_list_new[[1]], "best", ret = "threshold")
glm.pred_new <- ifelse(glm_probs_new[[1]] > thresh_new$threshold, 1, 0)
actual_new <- test_val$status_label_1yr

cm <- table(Predicted = glm.pred_new, Actual = actual_new)
cm

TP <- cm[2, 2]  
TN <- cm[1, 1]  
FP <- cm[2, 1]  
FN <- cm[1, 2]

accuracy  <- (TP + TN) / sum(cm)
precision <- TP / (TP + FP)
recall    <- TP / (TP + FN)
f1_score  <- 2 * precision * recall / (precision + recall)

cat("Accuracy :", round(accuracy, 3), "\n")
cat("Precision:", round(precision, 3), "\n")
cat("Recall   :", round(recall, 3), "\n")
cat("F1 Score :", round(f1_score, 3), "\n")
```

Our AUC has slightly improved, paired with other metrics such as Recall. However, precision has once again fallen, meaning even though we have a better AUC we are flagging more companies as bankrupt when they are not bankrupt given the removal of variables. 

```{r logit firth}
models_firth <- list()
for (i in names(formulas)) {
  f <- formulas[[i]]
  models_firth[[i]] <- logistf(f, data = train_val)
}

glm_probs_firth <- list()
for (i in names(models_firth)) {
  mod <- models_firth[[i]]
  glm_probs_firth[[i]] <- predict(mod, newdata = test_val, type = "response")
}

roc_list_firth <- list()
auc_values_firth <- numeric()

for (i in names(glm_probs_firth)) {
  roc_list_firth[[i]] <- roc(test_val$status_label_1yr, glm_probs_firth[[i]])
  auc_values_firth[i] <- auc(roc_list_firth[[i]])
}

# view AUC results
print(round(auc_values_firth, 3))

plot(roc_list_firth[[1]], col = 1, lwd = 2, main = "ROC Curves by Model")
for (j in 2:length(roc_list_firth)) {
  lines(roc_list_firth[[j]], col = j, lwd = 2)
}

legend("bottomright",
       legend = paste(names(roc_list_firth), "AUC =", round(auc_values_firth, 3)),
       col = 1:length(roc_list_firth), lwd = 2, bty = "n")

auc_df_firth <- data.frame(
  Model = c("Base", "Base + Lag1", "Base + Lag1 + Lag 2", "Base + All Lags"),
  AUC = c(auc_values_firth["base"], auc_values_firth["base_lag1"], auc_values_firth["base_lag1_2"], auc_values_firth["all"])
)

print(auc_df_firth)

best_index_firth <- which.max(auc_df_firth$AUC)
best_model_name_firth <- auc_df_firth$Model[best_index_firth]
print(best_model_name_firth)
```

Slightly better results regarding AUC when using Firth. Again, further study could delve into the mathematical reasoning behind such a slight increase.

```{r LASSO ENET RIDGE}
y_train <- as.integer(train_val$status_label_1yr == "failed")
x_train <- model.matrix(~ . - X.1 - X - status_label_1yr - status_label_2yr - status_label_3yr - company_name - year, data = train_val)
set.seed(1)
cvfit <- cv.glmnet(x_train,  train_val$status_label_1yr, family = "binomial", alpha = 1, type.measure = "auc")
plot(cvfit)
cvfit$lambda.min
cvfit$lambda.1se

cvfit_en <- cv.glmnet(x_train,  train_val$status_label_1yr, family = "binomial", alpha = 0.5, type.measure = "auc")
plot(cvfit_en)
cvfit_en$lambda.min
cvfit_en$lambda.1se

cvfit_ridge <- cv.glmnet(x_train,  train_val$status_label_1yr, family = "binomial", alpha = 0, type.measure = "auc")
plot(cvfit_ridge)
cvfit_ridge$lambda.min
cvfit_ridge$lambda.1se
```

LASSO, Elastic Net and Ridge AUC is slightly below previous tests, although all remain high. Future study could inspect these penalisation methods more in depth.

Highest AUC reported was base model, so proceed with that for "Results and Interpretation" section.

