---
title: "SVM test"
output: html_document
date: "2025-10-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(e1071)   
library(pROC)    
library(caret)   
library(ROSE)
library(dplyr)
library(smotefamily)
library(PRROC)
library(xgboost)
```

First we initialise the data
```{r}
df_train <- read.csv("C:\\Users\\railt\\DataScienceFanClub\\Data\\Val_traindata\\df_train_log_scaled_val.csv")
df_test <- read.csv("C:\\Users\\railt\\DataScienceFanClub\\Data\\Val_traindata\\df_test_log_scaled_val.csv")

df_train <- df_train[-c(1,2)]
df_test <- df_test[-c(1,2)]

feature_lag0 = c('X1','X2','X3','X4','X5','X6','X7','X8','X9','X10','X11','X12','X13','X14','X15','X17','X18')
feature_lag1 = c(feature_lag0, 'X1_1yr','X2_1yr','X3_1yr','X4_1yr','X5_1yr','X6_1yr','X7_1yr','X8_1yr','X9_1yr','X10_1yr','X11_1yr','X12_1yr','X13_1yr','X14_1yr','X15_1yr','X17_1yr','X18_1yr')
feature_lag2 = c(feature_lag1, 'X1_2yr','X2_2yr','X3_2yr','X4_2yr','X5_2yr','X6_2yr','X7_2yr','X8_2yr','X9_2yr','X10_2yr','X11_2yr','X12_2yr','X13_2yr','X14_2yr','X15_2yr','X17_2yr','X18_2yr')
feature_lag3 = c(feature_lag2, 'X1_3yr','X2_3yr','X3_3yr','X4_3yr','X5_3yr','X6_3yr','X7_3yr','X8_3yr','X9_3yr','X10_3yr','X11_3yr','X12_3yr','X13_3yr','X14_3yr','X15_3yr','X17_3yr','X18_3yr')

```

Initialise the data
```{r}

#no lag feature
x_train_nolag <- df_train[, feature_lag0]
x_test_nolag <- df_test[, feature_lag0]

#current year and include 1 lag year
x_train_lag1 <- df_train[, feature_lag1]
x_test_lag1 <- df_test[, feature_lag1]

#current year and include 1 and 2 lag years

x_train_lag2 <- df_train[, feature_lag2]
x_test_lag2 <- df_test[, feature_lag2]

#current year and include 1 and 2 and 3 lag years
x_train_lag3 <- df_train[, feature_lag3]
x_test_lag3 <- df_test[, feature_lag3]

y_train <- factor(df_train[,'status_label_1yr'], level = c('alive','failed'))
y_test <- factor(df_test[,'status_label_1yr'], level = c('alive','failed'))
```

```{r}
#smoting this data
set.seed(777) 
train_smote_nolag <- SMOTE(x_train_nolag, y_train)
train_smote_lag1 <- SMOTE(x_train_lag1, y_train)
train_smote_lag2 <- SMOTE(x_train_lag2, y_train)
train_smote_lag3 <- SMOTE(x_train_lag3, y_train)

new_df_nolag <- train_smote_nolag$data
new_df_lag1 <- train_smote_lag1$data
new_df_lag2 <- train_smote_lag2$data
new_df_lag3 <- train_smote_lag3$data

colnames(new_df_nolag)[ncol(new_df_nolag)] <- 'status_label_1yr'
colnames(new_df_lag1)[ncol(new_df_lag1)] <- 'status_label_1yr'
colnames(new_df_lag2)[ncol(new_df_lag2)] <- 'status_label_1yr'
colnames(new_df_lag3)[ncol(new_df_lag3)] <- 'status_label_1yr'

x_train_sm_nolag <- new_df_nolag[, -ncol(new_df_nolag)]
x_train_sm_lag1 <- new_df_lag1[, -ncol(new_df_lag1)]
x_train_sm_lag2 <- new_df_lag2[, -ncol(new_df_lag2)]
x_train_sm_lag3 <- new_df_lag3[, -ncol(new_df_lag3)]

y_train_sm_nolag <- factor(new_df_nolag$status_label_1yr)
y_train_sm_lag1 <- factor(new_df_lag1$status_label_1yr)
y_train_sm_lag2 <- factor(new_df_lag2$status_label_1yr)
y_train_sm_lag3 <- factor(new_df_lag3$status_label_1yr)
```
It takes approximately ~10-20 minutes to run each session of SVM with SMOTE upsampling. This is a conseuqnce of radial kernel SVM has approximately O(n^2) ~ O(n^3). Therefore for 4 values of c, and 4 values of possible gamma, if we were to run all possible combination of features i.e. no lag, included 1 lag , included 2 lag, included 3 lag, it would take about 16 sessions to complete the hyper parameter tuning. This amount to about 4.5 - 9 hours. This highlight a major downside with support vector machine method. In theory, the computational speed can be significantly reduce if linear kernel was chosen. However, linear kernel is a degenerative case of the radial kernel, and in most case produce a less accurage results (source). PCA can also possible reduce the computational time here but the algorithm scale linear with feature so it shouldn't affect the performance much.


In this section we will perform a parameter tuning. Given that it's extremely time consuming to perform tuning for a finer set of parameter. Here we simply perform a parameter tuning based on commonly use heuristic values. We first perform tuning for a model where the lag feature is not included
```{r}
cost_values  <- c(0.1, 1, 10, 100)
gamma_values <- c(2/ncol(x_train_sm_nolag), 1/ncol(x_train_sm_nolag), 0.1, 0.01)

results_nolag <- data.frame(cost = numeric(), gamma = numeric(),
                      Accuracy = numeric(), AUC = numeric(),
                      Sensitivity = numeric(), Specificity = numeric(), Confuse = numeric())

for (C in cost_values) {
  for (g in gamma_values) {

    svm_rbf <- svm(
      x = x_train_sm_nolag,
      y = y_train_sm_nolag,
      kernel = 'radial',
      cost   = C,
      gamma  = g,
      probability = TRUE
    )

    pred_oc <- predict(svm_rbf, x_test_nolag, decision.values = TRUE, probability = TRUE)
    pred    <- factor(pred_oc, levels = levels(y_test))

    # confusion matrix metrics
    cm <- confusionMatrix(pred, y_test, positive = "failed")

    # AUC from probabilities (alive)
    probs <- attr(pred_oc, "probabilities")[, "alive"]
    roc_obj <- roc(response = y_test,
                   predictor = probs)
    auc_val <- as.numeric(auc(roc_obj))
    
    print(auc_val)
    results_nolag <- rbind(results_nolag, data.frame(
      cost = C, gamma = g,
      Accuracy = as.numeric(cm$overall["Accuracy"]),
      AUC = auc_val,
      Sensitivity = as.numeric(cm$byClass["Sensitivity"]),
      Specificity = as.numeric(cm$byClass["Specificity"])
    ))
  }
}
```
tuning for model with 1 lag
```{r}
cost_values  <- c(0.1, 1, 10, 100)
gamma_values <- c(2/ncol(x_train_sm_lag1), 1/ncol(x_train_sm_lag1), 0.1, 0.01)

results_lag1 <- data.frame(cost = numeric(), gamma = numeric(),
                      Accuracy = numeric(), AUC = numeric(),
                      Sensitivity = numeric(), Specificity = numeric())

for (C in cost_values) {
  for (g in gamma_values) {

    svm_rbf <- svm(
      x = x_train_sm_lag1,
      y = y_train_sm_lag1,
      kernel = 'radial',
      cost   = C,
      gamma  = g,
      probability = TRUE
    )

    pred_oc <- predict(svm_rbf, x_test_lag1, decision.values = TRUE, probability = TRUE)
    pred    <- factor(pred_oc, levels = levels(y_test))

    # confusion matrix metrics
    cm <- confusionMatrix(pred, y_test, positive = "failed")

    probs <- attr(pred_oc, "probabilities")[, "alive"]
    roc_obj <- roc(response = y_test,
                   predictor = probs)
    auc_val <- as.numeric(auc(roc_obj))
    
    print(auc_val)
    results_lag1 <- rbind(results_lag1, data.frame(
      cost = C, gamma = g,
      Accuracy = as.numeric(cm$overall["Accuracy"]),
      AUC = auc_val,
      Sensitivity = as.numeric(cm$byClass["Sensitivity"]),
      Specificity = as.numeric(cm$byClass["Specificity"])
    ))
  }
}
```
model with 2yr lag
```{r}
cost_values  <- c(0.1, 1, 10, 100)
gamma_values <- c(2/ncol(x_train_sm_lag2), 1/ncol(x_train_sm_lag2), 0.1, 0.01)

results_lag2 <- data.frame(cost = numeric(), gamma = numeric(),
                      Accuracy = numeric(), AUC = numeric(),
                      Sensitivity = numeric(), Specificity = numeric())

for (C in cost_values) {
  for (g in gamma_values) {

    svm_rbf <- svm(
      x = x_train_sm_lag2,
      y = y_train_sm_lag2,
      kernel = 'radial',
      cost   = C,
      gamma  = g,
      probability = TRUE
    )

    pred_oc <- predict(svm_rbf, x_test_lag2, decision.values = TRUE, probability = TRUE)
    pred    <- factor(pred_oc, levels = levels(y_test))

    # confusion matrix metrics
    cm <- confusionMatrix(pred, y_test, positive = "failed")

    # AUC from probabilities (alive)
    probs <- attr(pred_oc, "probabilities")[, "alive"]
    roc_obj <- roc(response = y_test,
                   predictor = probs)
    auc_val <- as.numeric(auc(roc_obj))
    
    print(auc_val)
    results_lag2 <- rbind(results_lag2, data.frame(
      cost = C, gamma = g,
      Accuracy = as.numeric(cm$overall["Accuracy"]),
      AUC = auc_val,
      Sensitivity = as.numeric(cm$byClass["Sensitivity"]),
      Specificity = as.numeric(cm$byClass["Specificity"])
    ))
  }
}
```
model with 3yr lag
```{r}
cost_values  <- c(0.1, 1, 10, 100)
gamma_values <- c(2/ncol(x_train_sm_lag3), 1/ncol(x_train_sm_lag3), 0.1, 0.01)

results_lag3 <- data.frame(cost = numeric(), gamma = numeric(),
                      Accuracy = numeric(), AUC = numeric(),
                      Sensitivity = numeric(), Specificity = numeric())

for (C in cost_values) {
  for (g in gamma_values) {

    svm_rbf <- svm(
      x = x_train_sm_lag3,
      y = y_train_sm_lag3,
      kernel = 'radial',
      cost   = C,
      gamma  = g,
      probability = TRUE
    )

    pred_oc <- predict(svm_rbf, x_test_lag3, decision.values = TRUE, probability = TRUE)
    pred    <- factor(pred_oc, levels = levels(y_test))

    # confusion matrix metrics
    cm <- confusionMatrix(pred, y_test, positive = "failed")

    # AUC from probabilities (alive)
    probs <- attr(pred_oc, "probabilities")[, "alive"]
    roc_obj <- roc(response = y_test,
                   predictor = probs)
    auc_val <- as.numeric(auc(roc_obj))
    
    print(auc_val)
    results_lag3 <- rbind(results_lag3, data.frame(
      cost = C, gamma = g,
      Accuracy = as.numeric(cm$overall["Accuracy"]),
      AUC = auc_val,
      Sensitivity = as.numeric(cm$byClass["Sensitivity"]),
      Specificity = as.numeric(cm$byClass["Specificity"])
    ))
  }
}
```
Since it takes a great amount of time to train the model. In order to avoid having to retrain model everytime. I've decided to saved to result here and do an analysis in a seperated markdown file.
```{r}
write.csv(results_nolag, "C:\\Users\\railt\\DataScienceFanClub\\James\\results_nolag.csv")
write.csv(results_lag1, "C:\\Users\\railt\\DataScienceFanClub\\James\\results_lag1.csv")
write.csv(results_lag2, "C:\\Users\\railt\\DataScienceFanClub\\James\\results_lag2.csv")
write.csv(results_lag3, "C:\\Users\\railt\\DataScienceFanClub\\James\\results_lag3.csv")

```







