---
title: "Random Forest"
author: "Adlan"
date: "2025-10-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A random forest is an aggregation method which means we use predictions made from many decision trees to solve our binary classification problem.

One key advantage is that a single decision tree is sensitive to small changes in training data. By iterating over many decision trees that use different subsamples of the features and dataset, we reduce the importance of our training data and reduce overfitting — helping our model generalize as opposed to being specific to certain training data.

```{r}
library(tidyverse)
library(randomForest)
library(pROC)

# Function to train RF on a single year
train_rf_year <- function(train_file, test_file, year_suffix, seed=123) {
  # Load data
  train_data <- read.csv(train_file)
  test_data <- read.csv(test_file)
  
  # Target and predictor columns for this year
  target_col <- paste0("status_label_", year_suffix, "yr")
  train_data <- train_data %>% select(all_of(target_col), ends_with(paste0("_", year_suffix, "yr")))
  test_data <- test_data %>% select(all_of(target_col), ends_with(paste0("_", year_suffix, "yr")))
  
  # Convert target to factor
  train_data[[target_col]] <- as.factor(train_data[[target_col]])
  test_data[[target_col]] <- as.factor(test_data[[target_col]])
  
  train_data[[target_col]] <- droplevels(train_data[[target_col]])
  test_data[[target_col]] <- droplevels(test_data[[target_col]])
  
  # Train Random Forest
  set.seed(seed)
  rf_model <- randomForest(
    formula = as.formula(paste(target_col, "~ .")),
    data = train_data,
    ntree = 500,
    mtry = sqrt(ncol(train_data)-1),
    importance = TRUE,
    strata = train_data[[target_col]],
    sampsize = rep(min(table(train_data[[target_col]])), 2)
  )
  
  # Predictions and ROC
  prob_test <- predict(rf_model, test_data, type="prob")[,2]
  roc_obj <- roc(test_data[[target_col]], prob_test)
  auc_value <- auc(roc_obj)
  
  return(list(model=rf_model, roc=roc_obj, auc=auc_value))
}

# Train separate models for each year
rf_1yr <- train_rf_year("../data/train_scaled.csv", "../data/test_scaled.csv", 1)
rf_2yr <- train_rf_year("../data/train_scaled.csv", "../data/test_scaled.csv", 2)
rf_3yr <- train_rf_year("../data/train_scaled.csv", "../data/test_scaled.csv", 3)

# Compare AUCs
auc_df <- data.frame(
  Year = c("1-Year", "2-Year", "3-Year"),
  AUC = c(rf_1yr$auc, rf_2yr$auc, rf_3yr$auc)
)
print(auc_df)
```

```{r}
# Plot ROC curves for each year separately
plot(rf_1yr$roc, col="#1b9e77", lwd=2, main="ROC Curves by Year")
lines(rf_2yr$roc, col="#d95f02", lwd=2)
lines(rf_3yr$roc, col="#7570b3", lwd=2)
abline(a=0, b=1, lty=2, col="gray")
legend("bottomright", legend=paste(c("1-Year","2-Year","3-Year"), "AUC=", round(c(rf_1yr$auc, rf_2yr$auc, rf_3yr$auc), 3)),
       col=c("#1b9e77","#d95f02","#7570b3"), lwd=2, bty="n")
```

```{r}
# Feature importance for 1-year model (most recent year)
varImpPlot(rf_1yr$model, main="Variable Importance - 1-Year Features")
```








