---
title: "SVM test"
output: html_document
date: "2025-10-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
In this section 
```{r}
library(e1071)   
library(pROC)    
library(caret)   
library(ROSE)
library(dplyr)
library(smotefamily)
library(PRROC)
library(xgboost)
```

```{r}
df_train <- read.csv("C:\\Users\\railt\\DataScienceFanClub\\Data\\Val_traindata\\df_train_log_scaled_val.csv")
df_test <- read.csv("C:\\Users\\railt\\DataScienceFanClub\\Data\\Val_traindata\\df_test_log_scaled_val.csv")

df_train <- df_train[-c(1,2)]
df_test <- df_test[-c(1,2)]

feature_lag0 = c('X1','X2','X3','X4','X5','X6','X7','X8','X9','X10','X11','X12','X13','X14','X15','X17','X18')
feature_lag1 = c(feature_lag0, 'X1_1yr','X2_1yr','X3_1yr','X4_1yr','X5_1yr','X6_1yr','X7_1yr','X8_1yr','X9_1yr','X10_1yr','X11_1yr','X12_1yr','X13_1yr','X14_1yr','X15_1yr','X17_1yr','X18_1yr')
feature_lag2 = c(feature_lag1, 'X1_2yr','X2_2yr','X3_2yr','X4_2yr','X5_2yr','X6_2yr','X7_2yr','X8_2yr','X9_2yr','X10_2yr','X11_2yr','X12_2yr','X13_2yr','X14_2yr','X15_2yr','X17_2yr','X18_2yr')
feature_lag3 = c(feature_lag2, 'X1_3yr','X2_3yr','X3_3yr','X4_3yr','X5_3yr','X6_3yr','X7_3yr','X8_3yr','X9_3yr','X10_3yr','X11_3yr','X12_3yr','X13_3yr','X14_3yr','X15_3yr','X17_3yr','X18_3yr')

```

```{r}
x_train_present <- df_train[, feature_lag3]
x_test_present <- df_test[, feature_lag3]

y_train_present <- factor(df_train[,'status_label_1yr'])
y_test_present <- factor(df_test[,'status_label_1yr'])
```

```{r}
#smoting this data
set.seed(777) 
train_smote <- SMOTE(x_train_present,y_train_present)
new_df <- train_smote$data
colnames(new_df)[ncol(new_df)] <- 'status_label_1yr'
x_train_sm <- new_df[, -ncol(new_df)]
y_train_sm <- factor(new_df$status_label_1yr)
```
It takes approximately ~15-30 minutes to run each session of SVM with SMOTE upsampling. This is a conseuqnce of radial kernel SVM has approximately O(n^2) ~ O(n^3). Therefore for 3 values of c, and 3 values of possible gamma, if we were to run all possible combination of features i.e. no lag, included 1 lag , included 2 lag, included 3 lag, it would take about 36 sessions to complete the hyper parameter tuning. This amount to about 9 - 18 hours. This highlight a major downside with support vector machine method. In theory, the computational speed can be significantly reduce if linear kernel was chosen. However, linear kernel is a degenerative case of the radial kernel, and in most case produce a less accurage results (source). PCA can also possible reduce the computational time here but the algorithm scale linear with feature so it shouldn't affect the performance much.

```
svm_rbf <- svm(
  x = x_train_present,
  y = y_train_present,
  kernel = 'radial',
  cost=1,
  gamma = 1/ncol(x_train_sm)
)

pred_oc <- predict(svm_rbf, x_test_present, decision.values = TRUE)
pred <- factor(ifelse(pred_oc, "alive", "failed"), levels = levels(y_test_present))
print(confusionMatrix(pred, y_test_present, positive = "alive"))

```

```{r
pred_cls <- predict(svm_rbf, newdata = x_test_present)
print(confusionMatrix(pred_cls, y_test_present))
```

```{r}
cost_values  <- c(0.1, 1, 10, 100)
gamma_values <- c(2/ncol(x_train_sm), 1/ncol(x_train_sm), 0.1, 0.01)

results <- data.frame(cost = numeric(), gamma = numeric(),
                      Accuracy = numeric(), AUC = numeric(),
                      Sensitivity = numeric(), Specificity = numeric())

for (C in cost_values) {
  for (g in gamma_values) {

    svm_rbf <- svm(
      x = x_train_sm,
      y = y_train_sm,
      kernel = 'radial',
      cost   = C,
      gamma  = g,
      probability = TRUE
    )

    pred_oc <- predict(svm_rbf, x_test_present, decision.values = TRUE, probability = TRUE)
    pred    <- factor(pred_oc, levels = levels(y_test_present))

    # confusion matrix metrics
    cm <- confusionMatrix(pred, y_test_present, positive = "alive")

    # AUC from probabilities (alive)
    probs <- attr(pred_oc, "probabilities")[, "alive"]
    roc_obj <- roc(response = y_test_present,
                   predictor = probs,
                   levels = c("failed", "alive"),
                   direction = ">")
    auc_val <- as.numeric(auc(roc_obj))
    
    print(auc_val)
    results <- rbind(results, data.frame(
      cost = C, gamma = g,
      Accuracy = as.numeric(cm$overall["Accuracy"]),
      AUC = auc_val,
      Sensitivity = as.numeric(cm$byClass["Sensitivity"]),
      Specificity = as.numeric(cm$byClass["Specificity"])
    ))
  }
}

```


