---
title: "Rolling Window CV"
author: Tomi
output: html_document
date: "2025-11-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE}
library(xgboost)
library(dplyr)
library(pROC)
library(ggplot2)
library(caret)
```
Loading required datasets
```{r}
df <- read.csv("https://raw.githubusercontent.com/railtonjamesl-collab/DataScienceFanClub/refs/heads/main/Data/Final_traindata/df_train_full.csv")

df_test <- read.csv("https://raw.githubusercontent.com/railtonjamesl-collab/DataScienceFanClub/refs/heads/main/Data/Final_traindata/df_test.csv")
```

Features/label setup
```{r}
x <- paste0("X", 1:15)
base_features <- c(x, "X17", "X18")

lag_1yr_feats <- c(base_features, paste0(base_features, "_1yr"))
lag_2yr_feats <- c(lag_1yr_feats, paste0(base_features, "_2yr"))
lag_3yr_feats <- c(lag_2yr_feats, paste0(base_features, "_3yr"))

target <- "status_label_1yr"

first_year <- 2002
last_year <- 2013

# for rolling window CV

val_years <- c((first_year+1):last_year)
print(val_years)

set.seed(123)
```

Generating random parameters
```{r}
random_params <- function() {
  return(list(
    eta = runif(1, 0.01, 0.3),
    max_depth = sample(3:8, 1),
    subsample = runif(1, 0.7, 1.0),
    colsample_bytree = runif(1, 0.5, 1.0),
    min_child_weight = sample(1:5, 1),
    gamma = runif(1, 0, 5)
  ))
}
```

Defining the functions for training/evaluation
```{r cross-validation}
xgboost_rolling_cv <- function(features, parameters) {
  # Converting target variable to numeric
  df[[target]] <- as.numeric(df[[target]] == "failed")
  
  fold_aucs <- data.frame()
  
  company_col <- "company_name"
  
  for (val_year in val_years) {
    
    df_train <- df %>% filter(year < val_year)
    df_val <- df %>% filter(year == val_year)
    
    # Removing companies that appear on the validation set
    val_companies <- unique(df_val[[company_col]])
    df_train <- subset(df_train, !(df_train[[company_col]] %in% val_companies))
    
    dtrain <- xgb.DMatrix(data = as.matrix(df_train[, features]), label = df_train[[target]])
    dval <- xgb.DMatrix(data = as.matrix(df_val[, features]), label = df_val[[target]])
    
    # Handling class imbalance
    n_pos <- sum(df_train[[target]] == 1)
    n_neg <- sum(df_train[[target]] == 0)
    scale_pos_weight <- n_neg / n_pos
    print(scale_pos_weight)
    
    params <- list(
      objective = "binary:logistic",
      eta = parameters$eta,
      gamma = parameters$gamma,
      max_depth = parameters$max_depth,
      min_child_weight = parameters$min_child_weight,
      subsample = parameters$subsample,
      colsample_bytree = parameters$colsample_bytree,
      eval_metric = "auc",
      scale_pos_weight = scale_pos_weight
    )
    
    model <- xgb.train(
      params = params,
      data = dtrain,
      nrounds = 1000,
      watchlist = list(val = dval),
      print_every_n = 50,
      early_stopping_rounds = 30
    )
    
    preds <- predict(model, dval)
    auc_val <- as.numeric(auc(df_val[[target]], preds))
    best_iter <- model$best_iteration
    
    
    fold_aucs <- rbind(fold_aucs, data.frame(val_year = val_year, auc = auc_val, best_iter = best_iter))
  }
  
  return(list(
    mean_auc = mean(fold_aucs$auc),
    sd_auc = sd(fold_aucs$auc),
    fold_aucs = fold_aucs
  ))
}
```

```{r final eval function}
xgboost_train_test <- function(features, parameters, final_nrounds) {
  df[[target]] <- as.numeric(df[[target]] == "failed")
  df_test[[target]] <- as.numeric(df_test[[target]] == "failed")
  
  dtrain <- xgb.DMatrix(data = as.matrix(df[, features]), label = df[[target]])
  dtest <- xgb.DMatrix(data = as.matrix(df_test[, features]), label = df_test[[target]])
  
  # Handling class imbalance
  n_pos <- sum(df[[target]] == 1)
  n_neg <- sum(df[[target]] == 0)
  scale_pos_weight <- n_neg / n_pos
  
  print(scale_pos_weight)
  
  params <- list(
    objective = "binary:logistic",
    eta = parameters$eta,
    gamma = parameters$gamma,
    max_depth = parameters$max_depth,
    min_child_weight = 1,
    subsample = parameters$subsample,
    colsample_bytree = parameters$colsample_bytree,
    eval_metric = "auc",
    scale_pos_weight = scale_pos_weight
  )
    
  model <- xgb.train(
    params = params,
    data = dtrain,
    nrounds = final_nrounds
  )
  
  preds <- predict(model, dtest)
  auc_test <- auc(df_test[[target]], preds)
  roc_test <- roc(df_test[[target]], preds)
  return(list(auc = auc_test, roc = roc_test))
}
```

Random search for parameter tuning.
To find the optimal parameters, I ran 500 iterations of random search. I leave it at 50 here for lighter computation.
```{r}
results_base <- list()
results_1yr <- list()
results_2yr <- list()
results_3yr <- list()

for (i in 1:50) {
  parameters <- random_params()
  
  result0 <- xgboost_rolling_cv(base_features, parameters)
  result1 <- xgboost_rolling_cv(lag_1yr_feats, parameters)
  result2 <- xgboost_rolling_cv(lag_2yr_feats, parameters)
  result3 <- xgboost_rolling_cv(lag_3yr_feats, parameters)
  
  results_base[[i]] <- list(parameters = parameters, mean_auc = result0$mean_auc, sd_auc = result0$sd_auc, folds = result0$fold_aucs)
  
  results_1yr[[i]] <- list(parameters = parameters, mean_auc = result1$mean_auc, sd_auc = result1$sd_auc, folds = result1$fold_aucs)
  
  results_2yr[[i]] <- list(parameters = parameters, mean_auc = result2$mean_auc, sd_auc = result2$sd_auc, folds = result2$fold_aucs)
  
  results_3yr[[i]] <- list(parameters = parameters, mean_auc = result3$mean_auc, sd_auc = result3$sd_auc, folds = result3$fold_aucs)
}
```

Now we choose the optimal set of hyper parameters. They results are ranked by the mean AUC but the final choice also takes into account the standard deviation.
```{r base_features results set}
results_summary0 <- bind_rows(lapply(seq_along(results_base), function(i){
  cbind(
    iteration = i,
    as.data.frame(results_base[[i]]$parameters),
    mean_auc = results_base[[i]]$mean_auc,
    sd_auc = results_base[[i]]$sd_auc
  )
}))

# Display top results by mean AUC, showing the parameters used
results_summary0 <- results_summary0 %>% arrange(desc(mean_auc))
print(results_summary0[1:15, ])
```
Mean AUC distribution
```{r}
ggplot(results_summary0, aes(x = mean_auc)) +
  geom_histogram(bins = 22) +
  labs(title = "Distribution of mean AUC for different paramters", x = "Mean AUC", subtitle = "Base features")
```


AUC distribution across windows
```{r}
fold_aucs <- list()
for (i in 1:5) {
  fold_aucs[[i]] <- results_base[[i]]$folds$auc
}

fold_df <- bind_rows(lapply(fold_aucs, function(x) data.frame(auc = x)), .id = "params")

fold_df <- fold_df %>%
  group_by(params) %>%
  mutate(window = row_number()) %>%
  ungroup()

ggplot(
  data = fold_df,
  mapping = aes(x = window, y = auc, colour = params)
) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "AUC across windows", x = "Window", y = "AUC")
```

As seen above, the AUC scores vary a lot across windows.

```{r}
rank <- 1
idx <- results_summary0[rank, "iteration"]

auc_val_0 <- results_base[[idx]]$mean_auc
best_params_0 <- results_base[[idx]]$parameters
best_fold_aucs_0 <- results_base[[idx]]$folds

print(auc_val_0)
```


```{r base+1yr lag features results set}
results_summary1 <- bind_rows(lapply(seq_along(results_1yr), function(i){
  cbind(
    iteration = i,
    as.data.frame(results_1yr[[i]]$parameters),
    mean_auc = results_1yr[[i]]$mean_auc,
    sd_auc = results_1yr[[i]]$sd_auc
  )
}))

# Display top results by mean AUC, showing the parameters used
results_summary1 <- results_summary1 %>% arrange(desc(mean_auc))
print(results_summary1[1:15, ])
```

Mean AUC distribution
```{r}
ggplot(results_summary1, aes(x = mean_auc)) +
  geom_histogram(bins = 22) +
  labs(title = "Distribution of mean AUC for different paramters", x = "Mean AUC", subtitle = "Base + 1yr lag features")
```

```{r}
rank <- 2
idx <- results_summary1[rank, "iteration"]

auc_val_1 <- results_1yr[[idx]]$mean_auc
best_params_1 <- results_1yr[[idx]]$parameters
best_fold_aucs_1 <- results_1yr[[idx]]$folds

print(auc_val_1)
```

```{r base + 1yr + 2yr lag features results set}
results_summary2 <- bind_rows(lapply(seq_along(results_2yr), function(i){
  cbind(
    iteration = i,
    as.data.frame(results_2yr[[i]]$parameters),
    mean_auc = results_2yr[[i]]$mean_auc,
    sd_auc = results_2yr[[i]]$sd_auc
  )
}))

# Display top results by mean AUC, showing the parameters used
results_summary2 <- results_summary2 %>% arrange(desc(mean_auc))
print(results_summary2[1:15, ])
```
Mean AUC distribution
```{r}
ggplot(results_summary2, aes(x = mean_auc)) +
  geom_histogram(bins = 22) +
  labs(title = "Distribution of mean AUC for different paramters", x = "Mean AUC", subtitle = "Base + 1yr + 2yr lag features")
```

```{r}
rank <- 1 # similar to highest mean_auc but lower standard deviation
idx <- results_summary2[rank, "iteration"]

auc_val_2 <- results_2yr[[idx]]$mean_auc
best_params_2 <- results_2yr[[idx]]$parameters
best_fold_aucs_2 <- results_2yr[[idx]]$folds

print(auc_val_2)
```
```{r all features results set}
results_summary3 <- bind_rows(lapply(seq_along(results_3yr), function(i){
  cbind(
    iteration = i,
    as.data.frame(results_3yr[[i]]$parameters),
    mean_auc = results_3yr[[i]]$mean_auc,
    sd_auc = results_3yr[[i]]$sd_auc
  )
}))

# Display top results by mean AUC, showing the parameters used
results_summary3 <- results_summary3 %>% arrange(desc(mean_auc))
print(results_summary3[1:15, ])
```

Mean AUC distribution
```{r}
ggplot(results_summary2, aes(x = mean_auc)) +
  geom_histogram(bins = 22) +
  labs(title = "Distribution of mean AUC for different paramters", x = "Mean AUC", subtitle = "All features")
```

```{r}
rank <- 1
idx <- results_summary3[rank, "iteration"]

auc_val_3 <- results_3yr[[idx]]$mean_auc
best_params_3 <- results_3yr[[idx]]$parameters
best_fold_aucs_3 <- results_3yr[[idx]]$folds

print(auc_val_3)
```
Best parameters
```{r}
best_params <- list(best_params_2, iter = median(best_fold_aucs_2$best_iter))
print(best_params)
```

