---
title: "Logistic Regression and Further Study"
output: html_document
date: "2025-11-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r required libraries}
pkgs <- c("pROC", "car", "glmnet", "logistf")

for(p in pkgs){
    if (!require(p,character.only=TRUE)) install.packages(p)
}
```

```{r read}
file_path <- c("../Data/Val_traindata/df_train_log_scaled_val.csv", "../Data/Val_traindata/df_test_log_scaled_val.csv")
train_val <- read.csv(file_path[1])
test_val <- read.csv(file_path[2])
```

We will now convert the status year columns into numerical targets, a form of binary encoding, with 0 for "alive" and 1 for "failed".

```{r binary encoding}
cols <- c("status_label_1yr", "status_label_2yr", "status_label_3yr")

for (col in cols) {
  train_val[[col]] <- ifelse(train_val[[col]] == "failed", 1, 0)
  test_val[[col]]  <- ifelse(test_val[[col]] == "failed", 1, 0)
}
```

As explained in the group part of the report, we have introduced lag variables so it is important that we categorise them for easier use in the future.

```{r feature categorisation}
features <- paste0("X", 1:18)
features <- features[features !=  "X16"]
lag_1yr_feats <- paste0(features, "_1yr")
lag_2yr_feats <- paste0(features, "_2yr")
lag_3yr_feats <- paste0(features, "_3yr")

print(features)
```

We will now plot our simple logistic regression model, only considering base features and with respect to status_label_1yr.

```{r baseline test}
formula_base <- as.formula(paste("status_label_1yr ~", paste(features, collapse = " + ")))
model_base <- glm(formula_base, data = train_val, family = binomial())
vif(model_base)
summary(model_base)
```

There is high collinearity between variables. We will explore this more after our initial testing. 


```{r loop for models}
formulas <- list(
  base = as.formula(paste("status_label_1yr ~", paste(features, collapse = " + "))),
  base_lag1 = as.formula(paste("status_label_1yr ~", paste(c(features, lag_1yr_feats), collapse = " + "))),
  base_lag1_2 = as.formula(paste("status_label_1yr ~", paste(c(features, lag_1yr_feats, lag_2yr_feats), collapse = " + "))),
  all = as.formula(paste("status_label_1yr ~", paste(c(features, lag_1yr_feats, lag_2yr_feats, lag_3yr_feats), collapse = " + ")))
)

models <- list()
for (i in names(formulas)) {
  f <- formulas[[i]]
  models[[i]] <- glm(f, data = train_val, family = binomial())
}

for (i in names(models)) {
  cat("\nModel:", i, "\n")
  cat("AIC:", AIC(models[[i]]), "\n")
}
```

```{r af}
glm_probs <- list()
for (i in names(models)) {
  mod <- models[[i]]
  glm_probs[[i]] <- predict(mod, newdata = test_val, type = "response")
}

roc_list <- list()
auc_values <- numeric()

for (i in names(glm_probs)) {
  roc_list[[i]] <- roc(test_val$status_label_1yr, glm_probs[[i]])
  auc_values[i] <- auc(roc_list[[i]])
}

# view AUC results
print(round(auc_values, 3))

plot(roc_list[[1]], col = 1, lwd = 2, main = "ROC Curves by Model")
for (j in 2:length(roc_list)) {
  lines(roc_list[[j]], col = j, lwd = 2)
}

legend("bottomright",
       legend = paste(names(roc_list), "AUC =", round(auc_values, 3)),
       col = 1:length(roc_list), lwd = 2, bty = "n")

auc_df <- data.frame(
  Model = c("Base", "Base + Lag1", "Base + Lag1 + Lag 2", "Base + All Lags"),
  AUC = c(auc_values["base"], auc_values["base_lag1"], auc_values["base_lag1_2"], auc_values["all"])
)

print(auc_df)

best_index <- which.max(auc_df$AUC)
best_model_name <- auc_df$Model[best_index]
print(best_model_name)

thresh <- coords(roc_list[[1]], "best", ret = "threshold")
glm.pred <- ifelse(glm_probs[[1]] > thresh$threshold, 1, 0)
actual <- test_val$status_label_1yr

cm <- table(Predicted = glm.pred, Actual = actual)
cm

TP <- cm[2, 2]  
TN <- cm[1, 1]  
FP <- cm[2, 1]  
FN <- cm[1, 2]

accuracy  <- (TP + TN) / sum(cm)
precision <- TP / (TP + FP)
recall    <- TP / (TP + FN)
f1_score  <- 2 * precision * recall / (precision + recall)

cat("Accuracy :", round(accuracy, 3), "\n")
cat("Precision:", round(precision, 3), "\n")
cat("Recall   :", round(recall, 3), "\n")
cat("F1 Score :", round(f1_score, 3), "\n")
```

We see that the the base model has the highest AUC. We then calculate the confusion matrix and see that we have low precision but high recall. This is because our model is incorrectly flagging many companies as bankrupt when they do in fact not go bankrupt. However, the high AUC score of our Base Model does suggest that it can 

We now extend our findings by removing all highly correlated variables. We choose said variables by the correlation analysis done in the EDA section of the report.

```{r removal variables}
new_features <- features[-c(2,4,9,10)]
lag_1yr_newfeats <- paste0(new_features, "_1yr")
lag_2yr_newfeats <- paste0(new_features, "_2yr")
lag_3yr_newfeats <- paste0(new_features, "_3yr")

print(new_features)
```

```{r new feat}
formulas_new <- list(
  base_new = as.formula(paste("status_label_1yr ~", paste(new_features, collapse = " + "))),
  base_lag1_new = as.formula(paste("status_label_1yr ~", paste(c(new_features, lag_1yr_newfeats), collapse = " + "))),
  base_lag1_2_new = as.formula(paste("status_label_1yr ~", paste(c(new_features, lag_1yr_newfeats, lag_2yr_newfeats), collapse = " + "))),
  all_new = as.formula(paste("status_label_1yr ~", paste(c(new_features, lag_1yr_newfeats, lag_2yr_newfeats, lag_3yr_newfeats), collapse = " + ")))
)

models_new <- list()
for (i in names(formulas_new)) {
  f <- formulas_new[[i]]
  models_new[[i]] <- glm(f, data = train_val, family = binomial())
}

for (i in names(models_new)) {
  cat("\nModel:", i, "\n")
  cat("AIC:", AIC(models_new[[i]]), "\n")
}
```

```{r roc curves on new features}
glm_probs_new <- list()
for (i in names(models_new)) {
  mod <- models_new[[i]]
  glm_probs_new[[i]] <- predict(mod, newdata = test_val, type = "response")
}

roc_list_new <- list()
auc_values_new <- numeric()

# loop through all predicted probability vectors
for (i in names(glm_probs_new)) {
  roc_list_new[[i]] <- roc(test_val$status_label_1yr, glm_probs_new[[i]])
  auc_values_new[i] <- auc(roc_list_new[[i]])
}

# view AUC results
print(round(auc_values_new, 3))

# list roc plot
plot(roc_list_new[[1]], col = 1, lwd = 2, main = "ROC Curves by Model")
for (j in 2:length(roc_list_new)) {
  lines(roc_list_new[[j]], col = j, lwd = 2)
}

legend("bottomright",
       legend = paste(names(roc_list_new), "AUC =", round(auc_values_new, 3)),
       col = 1:length(roc_list_new), lwd = 2, bty = "n")

auc_df_new <- data.frame(
  Model = c("Base", "Base + Lag1", "Base + Lag1 + Lag2", "Base + All Lags"),
  AUC = c(auc_values_new["base_new"], auc_values_new["base_lag1_new"], auc_values_new["base_lag1_2_new"], auc_values_new["all_new"])
)

print(auc_df_new)

best_index_new <- which.max(auc_df_new$AUC)
best_model_name_new <- auc_df_new$Model[best_index_new]
print(best_model_name_new)
```

Our AUC has slightly improved. For further study, attempted to weight variables to counter act the imbalance class size in the dataset using the method outlined in King (2001) [1] to determine our weightings. Quickly discovered this method was not applicable. Attempt to use Firth logit [2], LASSO and Elastic Net [3].

```{r logit firth}
models_firth <- list()
for (i in names(formulas)) {
  f <- formulas[[i]]
  models_firth[[i]] <- logistf(f, data = train_val)
}

glm_probs_firth <- list()
for (i in names(models_firth)) {
  mod <- models_firth[[i]]
  glm_probs_firth[[i]] <- predict(mod, newdata = test_val, type = "response")
}

roc_list_firth <- list()
auc_values_firth <- numeric()

for (i in names(glm_probs_firth)) {
  roc_list_firth[[i]] <- roc(test_val$status_label_1yr, glm_probs_firth[[i]])
  auc_values_firth[i] <- auc(roc_list_firth[[i]])
}

# view AUC results
print(round(auc_values_firth, 3))

plot(roc_list_firth[[1]], col = 1, lwd = 2, main = "ROC Curves by Model")
for (j in 2:length(roc_list_firth)) {
  lines(roc_list_firth[[j]], col = j, lwd = 2)
}

legend("bottomright",
       legend = paste(names(roc_list_firth), "AUC =", round(auc_values_firth, 3)),
       col = 1:length(roc_list_firth), lwd = 2, bty = "n")

auc_df_firth <- data.frame(
  Model = c("Base", "Base + Lag1", "Base + Lag1 + Lag 2", "Base + All Lags"),
  AUC = c(auc_values_firth["base"], auc_values_firth["base_lag1"], auc_values_firth["base_lag1_2"], auc_values_firth["all"])
)

print(auc_df_firth)

best_index_firth <- which.max(auc_df_firth$AUC)
best_model_name_firth <- auc_df_firth$Model[best_index_firth]
print(best_model_name_firth)
```

```{r LASSO ENET RIDGE}
y_train <- as.integer(train_val$status_label_1yr == "failed")
x_train <- model.matrix(~ . - X.1 - X - status_label_1yr - status_label_2yr - status_label_3yr - company_name - year, data = train_val)
set.seed(1)
cvfit <- cv.glmnet(x_train,  train_val$status_label_1yr, family = "binomial", alpha = 1, type.measure = "auc")
plot(cvfit)
cvfit$lambda.min
cvfit$lambda.1se

cvfit_en <- cv.glmnet(x_train,  train_val$status_label_1yr, family = "binomial", alpha = 0.5, type.measure = "auc")
plot(cvfit_en)
cvfit_en$lambda.min
cvfit_en$lambda.1se

cvfit_ridge <- cv.glmnet(x_train,  train_val$status_label_1yr, family = "binomial", alpha = 0, type.measure = "auc")
plot(cvfit_ridge)
cvfit_ridge$lambda.min
cvfit_ridge$lambda.1se
```

LASSO and Elastic Net AUC is slightly below previous tests. For future study, it is possible to explore the reasons for this.

```{r final test}
file_path1 <- c("../Data/Final_traindata/df_train_log_scaled_full.csv", "../Data/Final_traindata/df_test_log_scaled.csv")
final_train <- read.csv(file_path1[1])
final_test <- read.csv(file_path1[2])

for (col in cols) {
  final_train[[col]] <- ifelse(final_train[[col]] == "failed", 1, 0)
  final_test[[col]]  <- ifelse(final_test[[col]] == "failed", 1, 0)
}

model_base_final <- glm(formulas[[1]], data = final_train, family = binomial())

glm_probs_basefinal = predict(model_base_final, newdata = final_test, type ="response")
roc_objbfinal <- roc(final_test$status_label_1yr, glm_probs_basefinal)

final_thresh <- coords(roc_objbfinal, "best", ret = "threshold")

glm.pred <- ifelse(glm_probs_basefinal > final_thresh$threshold, 1, 0)
actual <- final_test$status_label_1yr

cm <- table(Predicted = glm.pred, Actual = actual)
cm

TP <- cm[2, 2]  
TN <- cm[1, 1]  
FP <- cm[2, 1]  
FN <- cm[1, 2]

accuracy  <- (TP + TN) / sum(cm)
precision <- TP / (TP + FP)
recall    <- TP / (TP + FN)
f1_score  <- 2 * precision * recall / (precision + recall)

cat("Accuracy:", round(accuracy, 3), "\n")
cat("Precision:", round(precision, 3), "\n")
cat("Recall:", round(recall, 3), "\n")
cat("F1 Score:", round(f1_score, 3), "\n")

cat("Base Model AUC:", roc_objbfinal$auc, "\n")

plot(roc_list[[1]], col = 1, lwd = 2, main = "Final ROC Curves by Model")
legend("bottomright", legend = paste(("Base"), "AUC =", round(roc_objbfinal$auc, 3)), col = 1, lwd = 2, bty = "n")
```

```{r final test}
model_base_final_new <- glm(formulas_new[[1]], data = final_train, family = binomial())

glm_probs_basefinal_new = predict(model_base_final_new, newdata = final_test, type ="response")
roc_objbfinal_new <- roc(final_test$status_label_1yr, glm_probs_basefinal_new)

final_thresh_new <- coords(roc_objbfinal_new, "best", ret = "threshold")

glm.pred_new <- ifelse(glm_probs_basefinal_new > final_thresh_new$threshold, 1, 0)
actual_new <- final_test$status_label_1yr

cm <- table(Predicted = glm.pred_new, Actual = actual_new)
cm

TP <- cm[2, 2]  
TN <- cm[1, 1]  
FP <- cm[2, 1]  
FN <- cm[1, 2]

accuracy  <- (TP + TN) / sum(cm)
precision <- TP / (TP + FP)
recall    <- TP / (TP + FN)
f1_score  <- 2 * precision * recall / (precision + recall)

cat("Accuracy:", round(accuracy, 3), "\n")
cat("Precision:", round(precision, 3), "\n")
cat("Recall:", round(recall, 3), "\n")
cat("F1 Score:", round(f1_score, 3), "\n")

cat("Base Model AUC:", roc_objbfinal_new$auc, "\n")

plot(roc_list_new[[1]], col = 1, lwd = 2, main = "Final ROC Curves by Model")
legend("bottomright", legend = paste(("Base"), "AUC =", round(roc_objbfinal_new$auc, 3)), col = 1, lwd = 2, bty = "n")
```

References:
[1] https://gking.harvard.edu/sites/g/files/omnuum7116/files/gking/file

[3] https://glmnet.stanford.edu/articles/glmnet.html#logistic-regression-family-binomial